Mendelian randomization with genome-wide association studies data

### 2022-04-27 reply-to-review1/mr-review.Rmd
added math notation in latex

```{r, article-counts, message = FALSE, fig.cap= "Upward trend in article counts by year for keyword searches: 1. mendelian randomization and 2. genome-wide association study."}
library(magrittr)
dat <- readxl::read_xlsx(path = "../data/publication-counts.xlsx", sheet = "counts")
dat %>%
  tidyr::pivot_longer(!year) %>%
  dplyr::filter(year < 2022) %>%
  ggplot2::ggplot(ggplot2::aes(x = year, y = value, fill = name)) + 
  ggplot2::geom_bar(stat = "identity", 
                    position=ggplot2::position_dodge()) + 
  ggplot2::scale_fill_manual(labels = c("GWAS", "MR"), values=c('#999999','#E69F00')) + 
  ggplot2::ylab("Annual publication count") + 
  ggplot2::xlab("Year") + 
  #ggplot2::ggtitle("Published journal articles per year for Google Scholar key word searches: 1. Mendelian randomization and 2. Genome-wide association study") + 
  ggplot2::guides(fill = ggplot2::guide_legend("Search terms")) + 
  ggplot2::theme(text = ggplot2::element_text(size = 20))

@@

### Ratio of coefficients method
The ratio of coefficients method, also known as the Wald method, estimates the causal effect of the exposure X on the outcome Y by using a single SNP instrument [@wald_fitting_1940]. For a continuous outcome, the causal effect estimator,  is defined in Equation (3.1).

In Equation (3.1),  and  are the slope estimates from the regressions of the outcome and exposure, respectively, on the SNP instrument. Since the Wald method requires only the regression coefficients, it can be used with summary data. However, the Wald method doesn’t accommodate multiple SNP instruments, which limits its direct use in the GWAS setting as will be discussed later.

### Two-stage methods
A two-stage statistical model involves two regression models [@burgess_review_2017;@sawa_exact_1969]. For a continuous outcome, one may perform “two-stage least squares,” which involves two linear regressions. First, the exposure variable is regressed on the instrument (i.e., the SNP genotypes) (Equation (3.2)). We denote each subject’s exposure variable value as  in Equation (3.2), while  denotes an intercept term and  is the slope.  denotes the SNP instrument genotype for subject , while , the error term, is assumed to be independent among subjects and normally distributed with a shared common variance and mean zero.
The resulting fitted values for the exposure variable from Equation (3.2) are the independent variable in the second linear regression, where the outcome is the dependent variable. The causal effect estimate, then, is the coefficient  obtained from the second regression analysis (Equation (3.3)). Note also that the random errors in Equation (3.3) are assumed independent of those in Equation (3.2). Like the  in Equation (3.2), the  are assumed independent and identically distributed normal random errors.

@@

### Likelihood-based methods
Likelihood-based MR methods, unlike two-stage methods, provide maximum likelihood estimates with their many desirable properties [@fisher1925theory]. Limited information maximum likelihood from econometrics is the earliest approach for likelihood-based inference in MR [@davidson1993estimation;@hansen1996finite]. Limited information maximum likelihood with a single SNP instrument is modeled with two equations (Equations (3.4) and (3.5)), where the random errors follow a bivariate normal distribution [@burgess_review_2017].

Limited information maximum likelihood is sometimes called the maximum likelihood counterpart of two-stage least squares, and it yields the same causal estimate as two-stage least squares and the ratio method when used with a single SNP instrument. Additionally, the limited information maximum likelihood framework can accommodate  SNP instruments by replacing  with the sum .
One may also use Bayesian methods to obtain likelihood-based estimators [@kleibergen_bayesian_2003]. @burgess_improving_2012 examined a model that is similar to that from the limited information maximum likelihood framework (Equation (3.6)). The Bayesian model differs from the limited information maximum likelihood model in that the causal effect parameter,  represents the effect between the true means for the exposure and the outcome. In the limited information maximum likelihood model, the causal effect is that of the measured effect on the measured outcome.

For each subject , the exposure and outcome values come from a bivariate normal distribution. The mean of the exposure distribution is assumed linear in the SNP instrument, and the mean of the outcome distribution is linear in the mean exposure [@burgess_review_2017;@jones_choice_2012]. @burgess_improving_2012 demonstrated that the Bayesian model, with weak instruments, outperforms the limited information maximum likelihood model in terms of frequentist coverage levels.
### Semiparametric methods
Semiparametric instrumental variable methods, which feature parametric and nonparametric components, typically assume a parametric model connecting the exposure and outcome, but don’t impose distributional assumptions on the errors [@burgess_review_2017]. Compared to fully parametric models, semiparametric models often are more robust to model misspecification [@burgess_review_2017;@clarke_instrumental_2012]. We follow [@burgess_review_2017] by discussing three semiparametric strategies, generalized method of moments, continuous updating estimator, and G-estimation of structural mean models.
Generalized method of moments can be viewed as a more flexible form of two-stage least squares that handles heteroscedastic errors and nonlinearity in the two regressions [@burgess_review_2017]. Before specifying the model, we introduce notation.  is the conditional expectation of  if we forced  to take value  for every subject [@pearl2009causality].
The generalized method of moments equations, with a single SNP instrument, then, can be written as in Equation (3.7).

The GMM estimate is the value of the vector  that satisfies Equation (3.7), where . @palmer_instrumental_2011 developed numerical methods for obtaining estimates from Equations (3.7).
Another semiparametric estimation method is G-estimation of a structural mean model [@burgess_review_2017;@robins_new_1986;@greenland_estimating_2008].
We follow [@burgess_review_2017] by defining the potential outcome  as the outcome value that we would have observed had we set the exposure value  to . For example,  denotes the observed outcome had we set the exposure to zero instead of its observed value . The structural mean model for a continuous outcome is displayed in Equation (3.8).

The causal effect parameter is . [@burgess_review_2017] derives the estimating equations for , after noting that the conditional expectation,  is independent of  and reasoning that the causal effect is that value of  that yields zero covariance between  and  (Equations (3.9)).

Note that  indexes SNP instruments and ranges from  to  in Equation (3.9).

@@

One approach to integrating multiple SNP instruments into a MR framework is through calculation of allele scores [@burgess_use_2013]. @burgess_use_2013 calculate an unweighted score as the number of exposure-increasing alleles in the subject’s genotypes. They also calculate a weighted allele score by using the exposure effects as weights. For example, a subject with  copies of exposure-increasing alleles for SNP  has an unweighted allele score  and a weighted allele score . @burgess_use_2013 found that when weights are obtained from external data or from cross-validation or jackknife approaches applied to the analysis data, the allele score functions as a single instrumental variable and greatly diminishes the bias compared to that of the two-stage least squares estimator [@burgess_review_2017;@angrist1999jackknife].

@@

pleiotropic traits. For a collection of  SNPs, they choose the -vector that meets these criteria and maximizes the correlation with the exposure. To obtain their second estimator, @jiang_constrained_2019 apply a  penalty to constrain the number of SNP instruments that are given nonzero weights in the calculation of the new instrumental variable. They then compare their method with sisVIVE and allele score methods. @jiang_constrained_2019 find that both the smoothed version of their estimator and the allele score method are unbiased in their simulation settings.

@@

Figure 2: Two-sample MR with GWAS summary statistics methods publications density increases over time.
@burgess_mendelian_2013 termed their method “inverse variance weighting” (IVW) because each causal SNP’s effect is weighted by the inverse of the variance of the ratio estimator, and the overall causal effect is the sum of the weighted SNP causal effects.
Specifically, @burgess_mendelian_2013 combined causal effect ratio estimates from  independent SNPs by using Equation (5.1).

@bowden_consistent_2016, noting that consistency with IVW MR is not robust to invalid instrument use, developed a method for consistent MR analysis with
weighted median-based estimators, building on research from @han2008detecting. Its breakdown point is 50%, meaning that up to 50% of SNP instruments can be invalid while maintaining consistency in estimation [@hampel1971general]. Specifically, let  denote the  ordered ratio estimate statistic, from least to greatest. If an odd number, say , of SNP instruments is used, then the simple median estimator is defined as the  ordered ratio estimate. For an even number, say , number of SNP instruments, choose the midpoint between the two middle-ranking ordered ratio estimates, .
Due to the inefficiency of the simple median estimator, @bowden_consistent_2016 define the weighted median estimator. To construct this estimator, order the ratio estimates from least to greatest, as with the simple median estimator. The weighted median estimator, then is the median of a distribution defined by having  as the  percentile, where  is the weight assigned to the  ordered ratio estimate. In this setting, the simple median estimator is seen to be the weighted median estimator when all weights are equal.
@bowden_consistent_2016 also studied the penalized weighted median estimator. They defined the penalized weights, , where  is the p-value resulting from the comparison of  to a  distribution [@bowden_consistent_2016;@cochran1950comparison]. The penalized weighted median estimator leaves most variants unaffected, but downweights those with outlying Wald ratio estimates.
While the above MR methods all use independent SNP instruments, using independent instruments may not be ideal when there are multiple causal SNPs in linkage disequilibrium with each other. In this setting, discarding some of the causal SNPs may only capture a small proportion of trait variance explained by the exposure and lead to a power loss in MR [@burgess_review_2017;@burgess_mendelian_2013;@burgess_combining_2016;@burgess_bias_2011].  Consequently, many MR methods have been recently developed to model multiple correlated instruments. 
@burgess_combining_2016 developed methods for using correlated SNPs in IVW MR by inserting a covariance matrix into the standard IVW formulation. However, many MR methods use a set of pre-selected SNPs as instruments. Typically, these SNPs are selected to be statistically independent. This restriction to independent SNPs is needed for valid inference in methods like standard inverse variance weighting MR [@burgess_mendelian_2013].

## Assumption violations and Sensitivity Analysis
It is worthwhile to consider ways in which the standard MR assumptions may be violated and strategies for identifying such violations. Let’s first consider the exclusion restriction assumption (assumption 2 above). It states that the SNP instrument must affect the outcome only through the exposure. The causal diagram in Figure  illustrates a scenario that violates this assumption [@vanderweele_methodological_2014]. Bias in causal effect estimation would result. However, if the intermediate is measured and treated as the exposure, then an unbiased estimate of causal effect is possible, since the intermediate captures the two pathways, intermediate -> outcome and intermediate -> exposure -> outcome, through which the SNP instrument affects the outcome Y [@vanderweele_methodological_2014;@textor_graphical_2011;@textor_robust_2016].

Figure 3: Causal diagrams for MR. A. Scenario where the genetic variant affects an intermediate variable on the pathway to the exposure. Because the intermediate affects the outcome through a pathway that doesn’t involve exposure, this scenario violates the exclusion restriction assumption. B. Correlated horizontal pleiotropy occurs when the genetic variant affects the exposure, which in turn affects the outcome, and the genetic variant affects the unobserved confounder, which in turn affects the exposure and the outcome independently.

@@

where  designates confounders,  and  denote the exposure and outcome, respectively,  is a SNP instrument, and  and  are the zero-mean, normally distributed random errors.
@bowden_mendelian_2015 then write the outcome in terms of the SNP instrument  and the effects defined in Equation (5.2) and define the new parameter  (Equation (5.3)).

In Equation (5.3),  when SNP  is a valid instrument.
MR-Egger fits the regression model in Equation (5.4).

The notation in Equation (5.4) could be a little misleading. The independent and dependent variables are  and , respectively, while the regression parameters are  and .
The slope coefficient  estimates the causal effect of the exposure on the outcome. Under the assumption that the SNP-exposure relationship is independent of the SNP’s pleiotropic effects, Egger’s test offers a valid test of the null causal hypothesis and consistently estimates the causal effect, even if all SNPs are invalid instruments [@bowden_mendelian_2015].
In efforts to detect the presence of horizontal pleiotropy in two-sample MR studies with GWAS summary data, @greco_m_detecting_2015 developed the between-instrument heterogeneity  test. In a meta-analysis of MR Wald ratio estimates, one per SNP instrument, they calculate  in Equation (5.5).

In Equation (5.5),  is the inverse variance of the Wald estimator, and .
While the  test tends to be conservative in small sample sizes, its power increases with increasing sample size and increasing degree of pleiotropy [@greco_m_detecting_2015].

@hemani_orienting_2017, working in the setting of two-sample MR with summary GWAS data, recognized that measurement error in outcome and exposure can give misleading results with the inferred causal arrow in the wrong direction. To work around this issue, they extended MR with a method that infers the causal direction between two traits. Specifically, @hemani_orienting_2017 adapt Steiger’s Z test for a difference in correlations [@steiger1980tests] to assess which of the two traits in an analysis has the stronger correlation with the SNP instrument. That with the stronger correlation is treated as the exposure. @hemani_orienting_2017 applied their test in a study of DNA methylation and gene expression, where either direction of causality is plausible, and found that many methylation traits cause changes in gene expression.
@hartwig_robust_2017 developed a weighted mode-based MR causal effect estimation method for two-sample MR analysis with GWAS summary data. They first calculate Wald ratio estimates for every instrumental SNP. They then apply smoothing to the empirical distribution of ratio estimates. The mode of this smoothed distribution is the simple mode-based estimate of the causal effect [@hartwig_robust_2017]. The inverse variance-weighted mode-based estimate is obtained by weighting the empirical distribution of ratio estimates by the inverse variance of each estimate. @hartwig_robust_2017 compared their weighted and unweighted mode-based estimators with weighted and unweighted median-based estimators, MR Egger regression, and IVW. They found that their mode-based estimators demonstrated less bias and lower type I error rates than other estimators in some simulation settings. The mode-based estimators possessed lower statistical power to detect a causal effect compared to the IVW and weighted median-based methods, but their power exceeded that of MR Egger regression [@hartwig_robust_2017]. The mode-based estimators consistently estimated the causal effect when the mode across SNP instruments of the horizontal pleiotropy effects was zero [@hartwig_robust_2017]. In this manner, the mode-based estimators demonstrated a greater robustness to horizontal pleiotropy than did the other estimators.
@rees_robust_2019 approached the challenge of horizontal pleiotropy by developing three new methods for two-sample MR analysis with GWAS summary data: robust regression, penalized weights, and LASSO penalization. The first two can be viewed as modifications of MR-Egger and IVW methods. Together, they offer three strategies for downweighting or excluding variants with heterogeneous causal estimates. Recall that MR-Egger offers consistent causal effect estimates when there is no horizontal pleiotropy or when the horizontal pleiotropy effects adhere to the “Instrument Strength Independent of Direct Effect” (InSIDE) assumption [@bowden_mendelian_2015]. The InSIDE assumption is satisfied when there is no correlation between the pleiotropic SNP instrument-outcome effects and the SNP instrument-exposure effects. @rees_robust_2019 uses MM estimation, which is one technique for robust linear regression [@koller2011sharpening], to formulate an estimator with high breakdown point and high efficiency. The second estimation method, with penalized weights, is much like the penalized median estimator, from above, except that the second argument in the minimum function is , instead of the  above. Finally the LASSO-based method fashions an estimator by applying a penalty to the objective function from MR-Egger to get Equation (6.1) as the objective function.

@rees_robust_2019 studied two procedures for choosing : a heterogeneity stopping rule and a cross-validation rule. Taken together, these three methods offer additional contributions to a suite of sensitivity analysis tools that practitioners should consider.

>>>>

```{r, article-counts, message = FALSE, fig.cap= "Upward trend in article counts by year for keyword searches: 1. Mendelian randomization and 2. genome-wide association study."}
library(magrittr)
dat <- readxl::read_xlsx(path = "../data/publication-counts.xlsx", sheet = "counts")
dat %>%
  tidyr::pivot_longer(!year) %>%
  dplyr::filter(year < 2022) %>%
  ggplot2::ggplot(ggplot2::aes(x = year, y = value, fill = name)) + 
  ggplot2::geom_bar(stat = "identity", 
                    position=ggplot2::position_dodge()) + 
  ggplot2::scale_fill_manual(labels = c("GWAS", "MR"), values=c('#999999','#E69F00')) + 
  ggplot2::ylab("Annual publication count") + 
  ggplot2::xlab("Year") + 
  ggplot2::ggtitle("Published journal articles per year for Google Scholar key word searches: 1. Mendelian randomization and 2. Genome-wide association study") + 
  ggplot2::guides(fill = ggplot2::guide_legend("Search terms")) + 
  ggplot2::theme(text = ggplot2::element_text(size = 20))

@@

### Ratio of coefficients method
The ratio of coefficients method, also known as the Wald method, estimates the causal effect of the exposure X on the outcome Y by using a single SNP instrument [@wald_fitting_1940]. For a continuous outcome, the causal effect estimator, \@ref(eq:wald).

\begin{equation}
  \hat\gamma_{Wald} = \frac{\hat\beta_{YZ}}{\hat\beta_{XZ}}
  (\#eq:wald)
\end{equation}

In Equation \@ref(eq:wald), $\hat \beta_{YZ}$ and $\hat \beta_{XZ}$ are the slope estimates from the regressions of the outcome and exposure, respectively, on the SNP instrument. Since the Wald method requires only the regression coefficients, it can be used with summary data. However, the Wald method doesn’t accommodate multiple SNP instruments, which limits its direct use in the GWAS setting as will be discussed later.

### Two-stage methods
A two-stage statistical model involves two regression models [@burgess_review_2017;@sawa_exact_1969]. For a continuous outcome, one may perform “two-stage least squares,” which involves two linear regressions. First, the exposure variable is regressed on the instrument (i.e., the SNP genotypes) (Equation \@ref(eq:lm1)). We denote each subject’s exposure variable value as $x_i$ in Equation \@ref(eq:lm1), while $\alpha_0$ denotes an intercept term and $\alpha_1$ is the slope. $z_i$ denotes the SNP instrument genotype for subject $i$, while $\epsilon_i$, the error term, is assumed to be independent among subjects and normally distributed with a shared common variance and mean zero.

The resulting fitted values for the exposure variable from Equation \@ref(eq:lm1) are the independent variable in the second linear regression, where the outcome is the dependent variable. The causal effect estimate, then, is the coefficient  $\hat \beta_1$ obtained from the second regression analysis (Equation \@ref(eq:lm2)). Note also that the random errors in Equation \@ref(eq:lm2) are assumed independent of those in Equation \@ref(eq:lm1). Like the $\epsilon_i$ in Equation \@ref(eq:lm1), the $\tau_i$ are assumed independent and identically distributed normal random errors.

\begin{equation}
  x_i = \alpha_0 + \alpha_1 z_i + \epsilon_i
  (\#eq:lm1)
\end{equation}

\begin{equation}
  y_i = \beta_0 + \beta_1 \hat x_i + \tau_i
  (\#eq:lm2)
\end{equation}

@@

### Likelihood-based methods
Likelihood-based MR methods, unlike two-stage methods, provide maximum likelihood estimates with their many desirable properties [@fisher1925theory]. Limited information maximum likelihood from econometrics is the earliest approach for likelihood-based inference in MR [@davidson1993estimation;@hansen1996finite]. Limited information maximum likelihood with a single SNP instrument is modeled with two equations (Equations \@ref(eq:liml1) and \@ref(eq:liml2)), where the random errors follow a bivariate normal distribution [@burgess_review_2017].

\begin{equation}
  x_i = \alpha_0 + \alpha_1 z_{i1} + \epsilon_{Xi}
  (\#eq:liml1)
\end{equation}

\begin{equation}
  y_i = \beta_0 + \beta_1 x_i + \epsilon_{Yi}
  (\#eq:liml2)
\end{equation}

Limited information maximum likelihood is sometimes called the maximum likelihood counterpart of two-stage least squares, and it yields the same causal estimate as two-stage least squares and the ratio method when used with a single SNP instrument. Additionally, the limited information maximum likelihood framework can accommodate  more than one SNP instrument by replacing $\alpha_1z_{i1}$ with the sum $\sum_{k = 1}^K\alpha_k z_{ik}$.
One may also use Bayesian methods to obtain likelihood-based estimators [@kleibergen_bayesian_2003]. @burgess_improving_2012 examined a model that is similar to that from the limited information maximum likelihood framework (Equation \@ref(eq:b1)). The Bayesian model differs from the limited information maximum likelihood model in that the causal effect parameter, $\beta_1$ represents the effect between the true means for the exposure and the outcome. In the limited information maximum likelihood model, the causal effect is that of the measured effect on the measured outcome.

\begin{eqnarray}
\begin{pmatrix}
X_i \\
Y_i
\end{pmatrix} \sim N_2\left(\begin{pmatrix} \xi_i\\
\eta_i
\end{pmatrix}, \Sigma
\right)\nonumber \\
\xi_i = \alpha_0 + \alpha_1 z_{i1}\nonumber \\
\eta_i = \beta_0 + \beta_1 \xi_i
  (\#eq:b1)
\end{eqnarray}

For each subject $i$, the exposure and outcome values come from a bivariate normal distribution. The mean of the exposure distribution is assumed linear in the SNP instrument, and the mean of the outcome distribution is linear in the mean exposure [@burgess_review_2017;@jones_choice_2012]. @burgess_improving_2012 demonstrated that the Bayesian model, with weak instruments, outperforms the limited information maximum likelihood model in terms of frequentist coverage levels.

### Semiparametric methods
Semiparametric instrumental variable methods, which feature parametric and nonparametric components, typically assume a parametric model connecting the exposure and outcome, but don’t impose distributional assumptions on the errors [@burgess_review_2017]. Compared to fully parametric models, semiparametric models often are more robust to model misspecification [@burgess_review_2017;@clarke_instrumental_2012]. We follow [@burgess_review_2017] by discussing three semiparametric strategies, generalized method of moments, continuous updating estimator, and G-estimation of structural mean models.

Generalized method of moments can be viewed as a more flexible form of two-stage least squares that handles heteroscedastic errors and nonlinearity in the two regressions [@burgess_review_2017]. Before specifying the model, we introduce notation.  $\mathbb{E}(Y|\text{do}(X = x))$ is the conditional expectation of $Y$ if 
we forced $X$ to take value $x$ for every subject [@pearl2009causality].
The generalized method of moments equations, with a single SNP instrument, then, can be written as in Equation \@ref(eq:gmm).

\begin{eqnarray}
  \sum_i (y_i - f(x_i;\beta)) = 0\nonumber \\
  \sum_i z_i(y_i - f(x_i;\beta)) = 0
  (\#eq:gmm)
\end{eqnarray}

The GMM estimate is the value of the vector $\beta$ that satisfies Equation \@ref(eq:gmm), where $f(x_i; \beta) = \mathbb{E}(Y|\text{do}(X = x))$. @palmer_instrumental_2011 developed numerical methods for obtaining estimates from Equations \@ref(eq:gmm).

Another semiparametric estimation method is G-estimation of a structural mean model [@burgess_review_2017;@robins_new_1986;@greenland_estimating_2008].
We follow [@burgess_review_2017] by defining the potential outcome  as the outcome value that we would have observed had we set the exposure value  to . For example,  denotes the observed outcome had we set the exposure to zero instead of its observed value . The structural mean model for a continuous outcome is displayed in Equation (3.8).

$Y(x)$ as the outcome value that we would 
have observed had we set the exposure value $X$ to $x$. For example, $Y(0)$ denotes the observed outcome had 
we set the exposure to zero instead of its observed value $x$. The structural mean model for a continuous 
outcome is displayed in Equation \@ref(eq:smm).

\begin{equation}
\mathbb{E}(Y(x) - Y(0) | X = x, Z = z) = \beta_1 x
  (\#eq:smm)
\end{equation}

The causal effect parameter is $\beta_1$. @burgess_review_2017 derives the estimating equations for 
$\beta_1$, after noting that the conditional expectation, $\mathbb{E}(Y(0)| X = x, Z = z)$ is 
independent of $Z$ and reasoning that the causal effect is that value of $\beta_1$ that yields zero 
covariance between $Z$ and $\mathbb{E}(Y(0)| X = x, Z = z)$ (Equations \@ref(eq:smm1)).

\begin{equation}
\sum_i (z_{ik} - \bar z_{\cdot k})(y_i - \beta_1 x_i) = 0
  (\#eq:smm1)
\end{equation}

Note that $k$ indexes SNP instruments and ranges from $1$ to $K$ in Equation \@ref(eq:smm1).

@@

One approach to integrating multiple SNP instruments into a MR framework is through calculation of allele scores [@burgess_use_2013]. @burgess_use_2013 calculate an unweighted score as the number of exposure-increasing alleles in the subject’s genotypes. They also calculate a weighted allele score by using the exposure effects as weights. For example, a subject with $g_k$ copies of exposure-increasing alleles for SNP $k$ has an unweighted allele score $z = \sum_{k = 1}^K g_k$ and a weighted allele score $z = \sum_{k = 1}^K w_kg_k$. @burgess_use_2013 found that when weights are obtained from external data or from cross-validation or jackknife approaches applied to the analysis data, the allele score functions as a single instrumental variable and greatly diminishes the bias compared to that of the two-stage least squares estimator [@burgess_review_2017;@angrist1999jackknife].

@@

pleiotropic traits. For a collection of  SNPs, they choose the $K$-vector that meets these criteria and maximizes the correlation with the exposure. To obtain their second estimator, @jiang_constrained_2019 apply a $l_0$ penalty to constrain the number of SNP instruments that are given nonzero weights in the calculation of the new instrumental variable. They then compare their method with sisVIVE and allele score methods. @jiang_constrained_2019 find that both the smoothed version of their estimator and the allele score method are unbiased in their simulation settings.

@@

```{r, timeline, fig.cap = "Two-sample MR with GWAS summary statistics methods publications density increases over time."}
knitr::include_graphics("../figures/timeline.pdf")
```

@burgess_mendelian_2013 termed their method “inverse variance weighting” (IVW) because each causal SNP’s effect is weighted by the inverse of the variance of the ratio estimator, and the overall causal effect is the sum of the weighted SNP causal effects.
Specifically, @burgess_mendelian_2013 combined causal effect ratio estimates from  independent SNPs by using Equation \@ref(eq:ivw).

\begin{equation}
  \hat \beta_{IVW} = \frac{\sum_{j = 1}^M \hat\theta_j [Var(\hat\theta_j)]^{-1}}{\sum_{j = 1}^M [Var(\hat\theta_j)]^{-1}}
  (\#eq:ivw)
\end{equation}

@bowden_consistent_2016, noting that consistency with IVW MR is not robust to invalid instrument use, developed a method for consistent MR analysis with
weighted median-based estimators, building on research from @han2008detecting. Its breakdown point is 50%, meaning that up to 50% of SNP instruments can be invalid while maintaining consistency in estimation [@hampel1971general]. Specifically, let $\tilde \beta_j$ denote the $j^{th}$ ordered ratio estimate statistic, from least to greatest. If an odd number, say $2K = 1$, of SNP instruments is used, then the simple median estimator is defined as the $K + 1$ ordered ratio estimate. For an even number, say $2K$, number of SNP instruments, choose the midpoint between the two middle-ranking ordered ratio estimates, $\frac{\tilde \beta_K + \tilde \beta_{K + 1}}{2}$.
Due to the inefficiency of the simple median estimator, @bowden_consistent_2016 define the weighted median estimator. To construct this estimator, order the ratio estimates from least to greatest, as with the simple median estimator. The weighted median estimator, then is the median of a distribution defined by having $\tilde \beta_j$ as the $p_j = 100$ percentile, where $w_j$ is the weight assigned to the $j^{th}$ ordered ratio estimate. In this setting, the simple median estimator is seen to be the weighted median estimator when all weights are equal.
@bowden_consistent_2016 also studied the penalized weighted median estimator. They defined the 
penalized weights, $w_j^* = w_j' \times \min(1, 20q_j)$, where $q_j$ is the p-value resulting from 
the comparison of $Q_j = w_j'(\tilde\beta_j - \hat\beta_{IVW})^2$ to 
a $\chi^2_1$ distribution [@bowden_consistent_2016;@cochran1950comparison]. The penalized weighted median estimator leaves most variants unaffected, but downweights those with outlying Wald ratio estimates.
While the above MR methods all use independent SNP instruments, using independent instruments may not be ideal when there are multiple causal SNPs in linkage disequilibrium with each other. In this setting, discarding some of the causal SNPs may only capture a small proportion of trait variance explained by the exposure and lead to a power loss in MR [@burgess_review_2017;@burgess_mendelian_2013;@burgess_combining_2016;@burgess_bias_2011].  Consequently, many MR methods have been recently developed to model multiple correlated instruments. 
@burgess_combining_2016 developed methods for using correlated SNPs in IVW MR by inserting a covariance matrix into the standard IVW formulation. However, many MR methods use a set of pre-selected SNPs as instruments. Typically, these SNPs are selected to be statistically independent. This restriction to independent SNPs is needed for valid inference in methods like standard inverse variance weighting MR [@burgess_mendelian_2013].

## Assumption violations and Sensitivity Analysis

It is worthwhile to consider ways in which the standard MR assumptions may be violated and strategies for identifying such violations. Let’s first consider the exclusion restriction assumption (assumption 2 above). It states that the SNP instrument must affect the outcome only through the exposure. The causal diagram in Figure 3B illustrates a scenario that violates this assumption [@vanderweele_methodological_2014]. Bias in causal effect estimation would result. However, if the intermediate is measured and treated as the exposure, then an unbiased estimate of causal effect is possible, since the intermediate captures the two pathways, intermediate -> outcome and intermediate -> exposure -> outcome, through which the SNP instrument affects the outcome Y [@vanderweele_methodological_2014].

```{r, dags, fig.cap = "Causal diagrams for MR. A. Scenario where the genetic variant affects an exposure, which in turn affects an outcome. B. Scenario where the genetic variant affects an intermediate variable on the pathway to the exposure. Because the intermediate affects the outcome through a pathway that doesn’t involve exposure, this scenario violates the exclusion restriction assumption. C. Correlated horizontal pleiotropy occurs when the genetic variant affects the exposure, which in turn affects the outcome, and the genetic variant affects the unobserved confounder, which in turn affects the exposure and the outcome independently."}
knitr::include_graphics("../figures/dags.pdf")
```

@@

\begin{eqnarray}
X_i = \sum_{j=1}^J\gamma_jG_{ij} + U_i + \epsilon_{Xi}\nonumber \\
Y_i = \sum_{j=1}^J\alpha_jG_{ij} + \beta X_i + U_i + \epsilon_{Yi}
(\#eq:egger)
\end{eqnarray}

where $U$ designates confounders, $X$ and $Y$ denote the exposure and outcome, respectively, $G$ is a SNP 
instrument, and $\epsilon_X$ and $\epsilon_Y$ are the zero-mean, normally distributed random errors.

@bowden_mendelian_2015 then write the outcome in terms of the SNP instrument $j$ and 
the effects defined in Equation \@ref(eq:egger) and define the new parameter $\Gamma_j = \alpha_j + \beta\gamma_j$ (Equation \@ref(eq:egger1)).

\begin{eqnarray}
Y_i = (\alpha_j + \beta\gamma_j)G_{ij} +  \epsilon'_{Yij} \nonumber\\
= \Gamma_jG_{ij} + \epsilon'_{Yij}
(\#eq:egger1)
\end{eqnarray}

In Equation \@ref(eq:egger1), $\alpha_j = 0$ when SNP $j$ is a valid instrument.

MR-Egger fits the regression model in Equation \@ref(eq:egger2).

\begin{equation}
\hat\Gamma_j = \beta_{0E} + \beta_E \hat\gamma_j
(\#eq:egger2)
\end{equation}

The notation in Equation \@ref(eq:egger2) could be a little misleading. The independent
and dependent variables are $\hat\gamma_j$ and $\hat\Gamma_j$, respectively, while 
the regression parameters are $\beta_{0E}$ and $\beta_E$.

The slope coefficient $\hat\beta_E$ estimates the causal effect of the exposure on the outcome. 
Under the assumption that the SNP-exposure relationship is independent 
of the SNP's pleiotropic effects, Egger's test offers a valid test of 
the null causal hypothesis and consistently estimates the causal 
effect, even if all SNPs are invalid instruments [@bowden_mendelian_2015]. 
In efforts to detect the presence of horizontal pleiotropy in two-sample MR studies with GWAS summary data, @greco_m_detecting_2015 developed the between-instrument heterogeneity  test. In a meta-analysis 
of MR Wald ratio estimates, one per SNP instrument, they calculate $Q$ in Equation \@ref(eq:q).

\begin{equation}
Q = \sum_{k = 1}^K w_k(\hat\beta^{(k)}_{XY} - \mu_F)^2
(\#eq:q)
\end{equation}

In Equation \@ref(eq:q), $w_k$ is the inverse variance of the Wald estimator, and $\mu_F = \frac{\sum_{k = 1}^K w_k \hat\beta^{(k)}_{XY}}{\sum_{k = 1}^K w_k}$.

While the $Q$ test tends to be conservative in small sample sizes, its power increases with increasing 
sample size and increasing degree of pleiotropy [@greco_m_detecting_2015].

@hemani_orienting_2017, working in the setting of two-sample MR with summary GWAS data, recognized that measurement error in outcome and exposure can give misleading results with the inferred causal arrow in the wrong direction. To work around this issue, they extended MR with a method that infers the causal direction between two traits. Specifically, @hemani_orienting_2017 adapt Steiger’s Z test for a difference in correlations [@steiger1980tests] to assess which of the two traits in an analysis has the stronger correlation with the SNP instrument. That with the stronger correlation is treated as the exposure. @hemani_orienting_2017 applied their test in a study of DNA methylation and gene expression, where either direction of causality is plausible, and found that many methylation traits cause changes in gene expression.
@hartwig_robust_2017 developed a weighted mode-based MR causal effect estimation method for two-sample MR analysis with GWAS summary data. They first calculate Wald ratio estimates for every instrumental SNP. They then apply smoothing to the empirical distribution of ratio estimates. The mode of this smoothed distribution is the simple mode-based estimate of the causal effect [@hartwig_robust_2017]. The inverse variance-weighted mode-based estimate is obtained by weighting the empirical distribution of ratio estimates by the inverse variance of each estimate. @hartwig_robust_2017 compared their weighted and unweighted mode-based estimators with weighted and unweighted median-based estimators, MR Egger regression, and IVW. They found that their mode-based estimators demonstrated less bias and lower type I error rates than other estimators in some simulation settings. The mode-based estimators possessed lower statistical power to detect a causal effect compared to the IVW and weighted median-based methods, but their power exceeded that of MR Egger regression [@hartwig_robust_2017]. The mode-based estimators consistently estimated the causal effect when the mode across SNP instruments of the horizontal pleiotropy effects was zero [@hartwig_robust_2017]. In this manner, the mode-based estimators demonstrated a greater robustness to horizontal pleiotropy than did the other estimators.
@rees_robust_2019 approached the challenge of horizontal pleiotropy by developing three new methods for two-sample MR analysis with GWAS summary data: robust regression, penalized weights, and LASSO penalization. The first two can be viewed as modifications of MR-Egger and IVW methods. Together, they offer three strategies for downweighting or excluding variants with heterogeneous causal estimates. Recall that MR-Egger offers consistent causal effect estimates when there is no horizontal pleiotropy or when the horizontal pleiotropy effects adhere to the “Instrument Strength Independent of Direct Effect” (InSIDE) assumption [@bowden_mendelian_2015]. The InSIDE assumption is satisfied when there is no correlation between the pleiotropic SNP instrument-outcome effects and the SNP instrument-exposure effects. @rees_robust_2019 uses MM estimation, which is one technique for robust linear regression [@koller2011sharpening], to formulate an estimator with high breakdown point and high efficiency. The second estimation method, with penalized weights, is much like the penalized median estimator, from above, except that the second argument in the minimum function is , instead of the  above. Finally the LASSO-based method fashions an estimator by applying a penalty to the objective function from MR-Egger to get Equation \@ref(eq:lasso) as the objective function.

\begin{equation}
\sum_j [Var(\hat\beta_j)]^{-1}(\hat\beta_{Yj} - \theta_{0j} - \theta_1 \hat\beta_{1j})^2 + \lambda \sum_j |\theta_{0j}|
(\#eq:lasso)
\end{equation}

@rees_robust_2019 studied two procedures for choosing $\lambda$: a heterogeneity stopping rule 
and a cross-validation rule. Taken together, these three methods offer additional contributions to 
a suite of sensitivity analysis tools that practitioners should consider.
